On the propriety of restricted Boltzmann machines

### 2016-11-11 writing/draft.Rmd
abstrand and intro edits

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
date: "November, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has been largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability. We also describe the difficulties that arise in likelihood-based and Bayes fitting  of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

The data mining and machine learning communities have recently shown a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples]. A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]. In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical theory, and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

First, the models are often unsatisfactory as a conceptualization of how data are generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of the data generation process. In addition to such degeneracy, we find that RBMs can easily exhibit types of instability. In practice, this may seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing, @nguyen2014deep]. We investigate and these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.

@@

In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield unattractive models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.

For a RBM model with sufficient hidden variables, it has been shown that any distribution on the data can be approximated arbitrarily well [@le2008representational, @montufar2011refinements, and @montufar2011expressive]. The empirical distribution is the best fitting model to the observed cell data and so the RBM will seek to achieve this when possible. Thus, any fully principled fitting method based on the likelihood of a RBM with enough hidden variables will simply reproduce the (discrete) empirical distribution of most realistic training sets. That is, there is no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentially limiting the flexibility of the model.

This paper is structured as follows. Section \ref{background} formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of detecting it are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques intended to avoid fitted model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}. Additionally, there is an accompanying appendix with proofs of main results.

@@

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.

The RBM class of models can be thought to exhibit an analogous form of *near-degeneracy* as follows. When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
$$

@@

Additionally, S-unstable RBM model sequences are connected to degenerate model sequences placing all probability on a small portion of their sample spaces. To see this, define a modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_\nv, \nv \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.

@@

@@

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = LHS1)) +
  geom_contour(aes(x = r1, y = r2, z = LHS1), colour = "black", bins = 8) +
  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(frac(max(Q) - min(Q), H + V)), low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1) -> p.max_q

@@

## Bayesian model fitting
To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to near-degeneracy, instability, and uninterpretability. Motivated by the insights in Section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.

>>>>

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
date: "November, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs is largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model properties such as degeneracy, instability and uninterpretability. We also describe the difficulties that arise in likelihood-based and Bayes fitting  of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

The data mining and machine learning communities have recently shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples]. A RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]. In this paper, we provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical theory, and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

First, the models are often unsatisfactory as conceptualizations of how data are generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of a data generation process. In addition to such degeneracy, we find that RBMs can easily exhibit types of instability. In practice, this may seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing, @nguyen2014deep]. We investigate these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.

@@

In Section \ref{bayesian-model-fitting}, we compare three Bayesian techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of a RBM parameter space that yield unattractive models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.

For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational, @montufar2011refinements, and @montufar2011expressive]. The empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data and so principled RBM fitting will seek to produce this when possible. That is, any fully principled fitting method based on the likelihood for RBM with enough hidden variables will simply reproduce the (discrete) empirical distribution of a training set. Thus, there is no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentially limiting the flexibility of the fitted model.

This paper is structured as follows. Section \ref{background} formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques intended to avoid fitted model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}. Additionally, there is an accompanying appendix with proofs of main results.

@@

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. Conceptually, this makes sense because if a model is near-degenerate, then there are only a small number of elements of the sample space $\mathcal{X}$ that have positive probability, meaning the the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is only averaging that same number of values of $t( \boldsymbol x)$. Thus, this average value will not be pulled away from the boundary of $C$.

The RBM class of models can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space, $\mathcal{C}^{\nv + \nh}$. Using the method of @handcock2003assessing, when random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
$$

@@

Additionally, S-unstable RBM model sequences are connected to near-degeneracy of section \ref{near-degeneracy}, in which model sequences place all probability on a small portion of their sample spaces. To see this, define a modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_\nv, \nv \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.

@@

three_ways$Visibles <- factor(three_ways$Visibles, levels=rev(unique(three_ways$Visibles)))

@@

@@

## Bayesian model fitting
To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to near-degeneracy, instability, and uninterpretability. Motivated by the insights in Section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$. This is similar to an idea advocated by @hinton2010practical called *weight decay*, in which a penalty is added to the interaction terms in the model, $\boldsymbol \theta_{interaction}$, shrinking their magnitudes.
